---
title: "A4_STA457"
author: "Rafsaan Sanvir"
date: "2024-11-22"
output: word_document
---

```{r}
stock = scan("barrick.txt")
stock = ts(stock)
```

For adf test, H0: Time series is non-stationary
Ha: Time series is stationary 
```{r}
library(tseries) 
adf.test(stock)
adf.test(stock, k=2)
```
Since p > 0.05, we fail to reject H0 and conclude H0 that the time series is 
non-stationary. 

Fit ARIMA(0,1,1) and ARIMA(0,1,2) & extract residuals 
```{r}
ar_011 = arima(stock, order = c(0,1,1))
ar_012 = arima(stock, order = c(0,1,2))

resid_011= ar_011$residuals
resid_012= ar_012$residuals
```
To check if model fits data accurately and which of the 2 models to pick, we do
white noise tests on the residuals as well as use AIC to pick better model . 

```{r}
ar_011$aic
ar_012$aic
```
On the basis of AIC, we would prefer the ARIMA(0,1,2) model which has lower AIC 

H0: data is white noise 
HA: data is not white noise 
```{r}
Box.test(resid_011, lag = 10, fitdf = 1)
Box.test(resid_012, lag = 10, fitdf = 2)
```

We can see that both p value << 0.05 so we reject H0 and conclude that the data
does not resemble white noise. the p value for AR(0,1,2) is bigger so it is closer 
to white noise than the AR(0,1,1), making it a better pick as well. 

Let's pick AR(0,1,2) on the basis of AIC (IT HAS LOWER AI), and  use this in b 

1b) 
```{r}
residuals = resid_012
library(fGarch)

arch_1 = garchFit(residuals~garch(1, 0), data = residuals, trace = F)
arch_2 = garchFit(residuals~garch(2, 0), data = residuals, trace = F)
arch_3 = garchFit(residuals~garch(3, 0), data = residuals, trace = F)
arch_4 = garchFit(residuals~garch(4, 0), data = residuals, trace = F)
arch_5 = garchFit(residuals~garch(5, 0), data = residuals, trace = F)
```
This fits the model et = mu + sigmat * vt. If the model is approprite, then the 
residuals should look like independent observations from the density f, where f 
is the density function of vt. By default it is usually N(0,1) gaussian. 
We can look at other tests as well. 


Let's look at ARCH(1) and ARCH(5) models 

```{r}
summary(arch_1)
summary(arch_5)
```

The jaque bera test rejects the hypothesis that the residuals are close to gaussian 
white noise for both models as it is essentially 0. 
The Ljung-box test for R^2 is rejecting the fact that the squared residuals are 
white noise.

Now we select model based on AIC/BIC. Our best model using AIC would be the ARCH(5) 
model, as we can see a trend where AIC decreases as m increases from 1 to 5. Just 
for cleanliness, I will not plot the other 3 models but I checked. 

c) 
```{r}
garch_1 = garchFit(residuals~garch(1, 1), data = residuals, trace = F)
garch_2 = garchFit(residuals~garch(1, 2), data = residuals, trace = F)
garch_3 = garchFit(residuals~garch(1, 3), data = residuals, trace = F)

summary(garch_1)
summary(garch_2)
summary(garch_3)
```

We can see that all 3 models are an improvement from the ARCH(5) models, because 
the ljung box test p value is fairly large. 
H0: data resembles white noise 
Ha: data does not resemble white noise. 

So here, we fail to  reject H0 and conclude H0 that the residuals squared resemble
white noise, which is exactly what we wanted! 

the lowest AIC and BIC is for the GARCH(1,2) model which would be most approprite 
to model the residuals of the ARIMA(0,1,2) model. 

__________________________________________________________________________________


2d) I will use the code from assignment 1 and same to recover as. I will also 
include my own comments to understand this method better. 
```{r}
M = 2048  #no of frequency points
lambda = 100  #smoothing parameter?? 
omega = seq(0, 1, length.out = M) #frequency 

fx = 4 * (omega - 0.5)^2  #sdf of X_t
fxy = (omega - 0.5)^4  #cross-spec

# Compute coefficients a_s using FFT
gamma = fxy / fx  #transfer function
gamma_fft <- fft(gamma)  
as <- Re(fft(gamma, inverse = TRUE)) / M 
as_centered <- c(as[(M - 20 + 1):M], as[1:21])
s <- c(-20:20)  # Range of s values

# Plot
plot(s, as_centered, ylab = "as", main = "Coefficients a_s", 
     type = "h", lwd = 6, col = "red") #plot as from -20 to 20
abline(h = 0, col = "gray", lwd = 2)  
grid() 
```





